Faithful Model Explanations through\\ Energy-Constrained Conformal Counterfactuals

### 2023-12-16 paper/partials/appendix.tex
just using separate file

\import{contents/}{table_ebm_params.tex}

@@

\import{contents/}{table_params.tex}

\import{contents/}{table_perf.tex}

\import{contents/}{table_gen_params.tex}

@@

\import{contents/}{table-linearly-separable.tex}

\import{contents/}{table-circles.tex}

\import{contents/}{table-moons.tex}

\import{contents/}{table-california-housing.tex}

\import{contents/}{table-gmsc.tex}

\import{contents/}{table-german-credit.tex}

\import{contents/}{table-mnist.tex}

\import{contents/}{table-fashion-mnist.tex}

\import{contents/}{table-linearly-separable-valid.tex}

\import{contents/}{table-circles-valid.tex}

\import{contents/}{table-moons-valid.tex}

\import{contents/}{table-california-housing-valid.tex}

\import{contents/}{table-gmsc-valid.tex}

\import{contents/}{table-german-credit-valid.tex}

\import{contents/}{table-mnist-valid.tex}

\import{contents/}{table-fashion-mnist-valid.tex}

>>>>

\import{../contents/}{table_ebm_params.tex}

@@

\import{../contents/}{table_params.tex}

\import{../contents/}{table_perf.tex}

\import{../contents/}{table_gen_params.tex}

@@

\import{../contents/}{table-linearly-separable.tex}

\import{../contents/}{table-circles.tex}

\import{../contents/}{table-moons.tex}

\import{../contents/}{table-california-housing.tex}

\import{../contents/}{table-gmsc.tex}

\import{../contents/}{table-german-credit.tex}

\import{../contents/}{table-mnist.tex}

\import{../contents/}{table-fashion-mnist.tex}

\import{../contents/}{table-linearly-separable-valid.tex}

\import{../contents/}{table-circles-valid.tex}

\import{../contents/}{table-moons-valid.tex}

\import{../contents/}{table-california-housing-valid.tex}

\import{../contents/}{table-gmsc-valid.tex}

\import{../contents/}{table-german-credit-valid.tex}

\import{../contents/}{table-mnist-valid.tex}

\import{../contents/}{table-fashion-mnist-valid.tex}

### 2023-12-16 paper/partials/body.tex
just using separate file

%File: formatting-instructions-latex-2024.tex
%release 2024.0
\documentclass[letterpaper]{article} % DO NOT CHANGE THIS
\usepackage{aaai24}  % DO NOT CHANGE THIS
\usepackage{times}  % DO NOT CHANGE THIS
\usepackage{helvet}  % DO NOT CHANGE THIS
\usepackage{courier}  % DO NOT CHANGE THIS
\usepackage[hyphens]{url}  % DO NOT CHANGE THIS
\usepackage{graphicx} % DO NOT CHANGE THIS
\urlstyle{rm} % DO NOT CHANGE THIS
\def\UrlFont{\rm}  % DO NOT CHANGE THIS
\usepackage{natbib}  % DO NOT CHANGE THIS AND DO NOT ADD ANY OPTIONS TO IT
\usepackage{caption} % DO NOT CHANGE THIS AND DO NOT ADD ANY OPTIONS TO IT
\frenchspacing  % DO NOT CHANGE THIS
\setlength{\pdfpagewidth}{8.5in}  % DO NOT CHANGE THIS
\setlength{\pdfpageheight}{11in}  % DO NOT CHANGE THIS
%
% These are recommended to typeset algorithms but not required. See the subsubsection on algorithms. Remove them if you don't have algorithms in your paper.
\usepackage{algorithm}
% \usepackage{algorithmic}

%
% These are are recommended to typeset listings but not required. See the subsubsection on listing. Remove this block if you don't have listings in your paper.
\usepackage{newfloat}
\usepackage{listings}
\DeclareCaptionStyle{ruled}{labelfont=normalfont,labelsep=colon,strut=off} % DO NOT CHANGE THIS
\lstset{%
	basicstyle={\footnotesize\ttfamily},% footnotesize acceptable for monospace
	numbers=left,numberstyle=\footnotesize,xleftmargin=2em,% show line numbers, remove this entire line if you don't want the numbers.
	aboveskip=0pt,belowskip=0pt,%
	showstringspaces=false,tabsize=2,breaklines=true}
\floatstyle{ruled}
\newfloat{listing}{tb}{lst}{}
\floatname{listing}{Listing}
%
% Keep the \pdfinfo as shown here. There's no need
% for you to add the /Title and /Author tags.
\pdfinfo{
/TemplateVersion (2024.1)
}

% DISALLOWED PACKAGES
% \usepackage{authblk} -- This package is specifically forbidden
% \usepackage{balance} -- This package is specifically forbidden
% \usepackage{color (if used in text)
% \usepackage{CJK} -- This package is specifically forbidden
% \usepackage{float} -- This package is specifically forbidden
% \usepackage{flushend} -- This package is specifically forbidden
% \usepackage{fontenc} -- This package is specifically forbidden
% \usepackage{fullpage} -- This package is specifically forbidden
% \usepackage{geometry} -- This package is specifically forbidden
% \usepackage{grffile} -- This package is specifically forbidden
% \usepackage{hyperref} -- This package is specifically forbidden
% \usepackage{navigator} -- This package is specifically forbidden
% (or any other package that embeds links such as navigator or hyperref)
% \indentfirst} -- This package is specifically forbidden
% \layout} -- This package is specifically forbidden
% \multicol} -- This package is specifically forbidden
% \nameref} -- This package is specifically forbidden
% \usepackage{savetrees} -- This package is specifically forbidden
% \usepackage{setspace} -- This package is specifically forbidden
% \usepackage{stfloats} -- This package is specifically forbidden
% \usepackage{tabu} -- This package is specifically forbidden
% \usepackage{titlesec} -- This package is specifically forbidden
% \usepackage{tocbibind} -- This package is specifically forbidden
% \usepackage{ulem} -- This package is specifically forbidden
% \usepackage{wrapfig} -- This package is specifically forbidden
% DISALLOWED COMMANDS
% \nocopyright -- Your paper will not be published if you use this command
% \addtolength -- This command may not be used
% \balance -- This command may not be used
% \baselinestretch -- Your paper will not be published if you use this command
% \clearpage -- No page breaks of any kind may be used for the final version of your paper
% \columnsep -- This command may not be used
% \newpage -- No page breaks of any kind may be used for the final version of your paper
% \pagebreak -- No page breaks of any kind may be used for the final version of your paperr
% \pagestyle -- This command may not be used
% \tiny -- This is not an acceptable font size.
% \vspace{- -- No negative value may be used in proximity of a caption, figure, table, section, subsection, subsubsection, or reference
% \vskip{- -- No negative value may be used to alter spacing above or below a caption, figure, table, section, subsection, subsubsection, or reference

% Additional Packages:
\usepackage{amsfonts}       % blackboard math symbols
\usepackage{amsmath}
\usepackage{amsthm}
\usepackage{caption}
\usepackage{graphicx}
\usepackage{algpseudocode}
\usepackage{import}
\usepackage{booktabs}
\usepackage{longtable}
\usepackage{array}
\usepackage{multirow}
\usepackage{placeins}

% Numbered Environments:
\newtheorem{definition}{Definition}[section]
\newtheorem{question}{Research Question}[section]

\setcounter{secnumdepth}{1} %May be changed to 1 or 2 if section numbers are desired.

% The file aaai24.sty is the style file for AAAI Press
% proceedings, working notes, and technical reports.
%

% Title

% Your title must be in mixed case, not sentence case.
% That means all verbs (including short verbs like be, is, using,and go),
% nouns, adverbs, adjectives should be capitalized, including both words in hyphenated terms, while
% articles, conjunctions, and prepositions are lower case unless they
% directly follow a colon or long dash

\title{Faithful Model Explanations through\\
Energy-Constrained Conformal Counterfactuals}
\author{
    %Authors
    % All authors must be in the same font size and format.
    Patrick Altmeyer\textsuperscript{\rm 1},
    Mojtaba Farmanbar\textsuperscript{\rm 2},
    Arie van Deursen\textsuperscript{\rm 1},
    Cynthia C. S. Liem\textsuperscript{\rm 1}
}
\affiliations{
    %Afiliations
    \textsuperscript{\rm 1}Delft University of Technology,\\
    Faculty of Electrical Engineering, Mathematics and Computer Science\\
    \textsuperscript{\rm 2}ING Bank\\
    p.altmeyer@tudelft.nl
%
% See more examples next
}

\begin{document}

\maketitle

\begin{abstract}
    Counterfactual explanations offer an intuitive and straightforward way to explain black-box models and offer algorithmic recourse to individuals. To address the need for plausible explanations, existing work has primarily relied on surrogate models to learn how the input data is distributed. This effectively reallocates the task of learning realistic explanations for the data from the model itself to the surrogate. Consequently, the generated explanations may seem plausible to humans but need not necessarily describe the behaviour of the black-box model faithfully. We formalize this notion of faithfulness through the introduction of a tailored evaluation metric and propose a novel algorithmic framework for generating \textbf{E}nergy-\textbf{C}onstrained \textbf{C}onformal \textbf{Co}unterfactuals that are only as plausible as the model permits. Through extensive empirical studies, we demonstrate that \textit{ECCCo} reconciles the need for faithfulness and plausibility. In particular, we show that for models with gradient access, it is possible to achieve state-of-the-art performance without the need for surrogate models. To do so, our framework relies solely on properties defining the black-box model itself by leveraging recent advances in energy-based modelling and conformal prediction. To our knowledge, this is the first venture in this direction for generating faithful counterfactual explanations. Thus, we anticipate that \textit{ECCCo} can serve as a baseline for future research. We believe that our work opens avenues for researchers and practitioners seeking tools to better distinguish trustworthy from unreliable models.
\end{abstract}

@@

\begin{figure}
  \centering
  \includegraphics[width=0.8\linewidth]{figures/mnist_motivation.png}
  \caption{Counterfactuals for turning a 9 (nine) into a 7 (seven): original image (left), then the counterfactuals generated using \textit{Wachter}, \textit{Schut} and \textit{REVISE}.}\label{fig:motiv}
\end{figure}

@@

\begin{figure*}
  \centering
  \includegraphics[width=0.75\linewidth]{figures/poc_gradient_fields.png}
  \caption{Gradient fields and counterfactual paths for different generators. The objective is to generate a counterfactual in the blue class for a sample from the orange class. Bright yellow stars indicate conditional samples generated through SGLD. The underlying classifier is a Joint Energy Model.}\label{fig:poc}
\end{figure*}

@@

\begin{figure}
  \centering
  \includegraphics[width=1.0\linewidth]{figures/mnist_benchmark.png}
  \caption{Counterfactuals for turning a 3 into a 5: factual (left), then the counterfactuals generated by \textit{ECCCo}, \textit{ECCCo+}, \textit{REVISE}, \textit{Schut} and \textit{Wachter}.}\label{fig:mnist-bmk}
\end{figure}

@@

\bibliography{aaai24,bib}

\end{document}

>>>>

@@

\begin{figure}
  \centering
  \includegraphics[width=0.8\linewidth]{../camera-ready/figures/mnist_motivation.png}
  \caption{Counterfactuals for turning a 9 (nine) into a 7 (seven): original image (left), then the counterfactuals generated using \textit{Wachter}, \textit{Schut} and \textit{REVISE}.}\label{fig:motiv}
\end{figure}

@@

\begin{figure*}
  \centering
  \includegraphics[width=0.75\linewidth]{../camera-ready/figures/poc_gradient_fields.png}
  \caption{Gradient fields and counterfactual paths for different generators. The objective is to generate a counterfactual in the blue class for a sample from the orange class. Bright yellow stars indicate conditional samples generated through SGLD. The underlying classifier is a Joint Energy Model.}\label{fig:poc}
\end{figure*}

@@

\begin{figure}
  \centering
  \includegraphics[width=1.0\linewidth]{../camera-ready/figures/mnist_benchmark.png}
  \caption{Counterfactuals for turning a 3 into a 5: factual (left), then the counterfactuals generated by \textit{ECCCo}, \textit{ECCCo+}, \textit{REVISE}, \textit{Schut} and \textit{Wachter}.}\label{fig:mnist-bmk}
\end{figure}

@@
